---
title: "DESeq2"
author: "PNP"
date: "2026-05-29"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
params:
  in_dir: "example_multiple_conditions/input"
  out_dir: "example_multiple_conditions/output"
header-includes:
  - \let\oldtableofcontents\tableofcontents
  - \renewcommand{\tableofcontents}{\oldtableofcontents\clearpage}
---

# Initial setup

```{r Rmd setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/project/def-mouellet/Scripts_MOU/PNP/alliancecan/deseq2/")
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.align = "center")
```

```{r Set working directory}
# Set working directory
setwd("/project/def-mouellet/Scripts_MOU/PNP/alliancecan/deseq2/")
getwd()
```

```{r Activate renv}
# Activate environment
renv::activate()
```

```{r Load libraries}
# Load required packages
library(DESeq2)
library(ggplot2)
library(tidyverse)
library(limma)
library(pheatmap)
library(RColorBrewer)
library(ggrepel)
library(DEGreport)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggnewscale)
library(pathview)

# Set seed
set.seed(42)
```

```{r Input dir}
# Input directory
in_dir <- params$in_dir
print(in_dir)
dir(in_dir)
```

```{r Output dir}
# Output directory
out_dir <- params$out_dir

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

print(out_dir)
```

# Load data

## Count data

```{r}
# Count data
data <- read.table(
  file.path(in_dir, "Mov10_full_counts.txt"), 
  sep = "\t", header = T, row.names = 1
  )

head(data)
```

```{r}
str(data)
```

```{r}
# If count data is not integer, round to integer
# data <- data %>%
#   mutate(across(where(is.numeric), ~ as.integer(round(.))))
```

```{r}
# To remove samples
# data <- data %>% 
#   dplyr::select(-c(Irrel_kd_3))
```

```{r}
data %>%
  dim()
```

## Col data / Sample info / Metadata

```{r}
# Col data
meta <- read.table(
  file.path(in_dir, "Mov10_full_meta.txt"),
  sep = "\t", header = T, row.names = 1
  )

meta
```

```{r}
# Convert to factor
meta[] <- lapply(meta, as.factor) # all columns
str(meta)
```

```{r}
# To remove samples
# meta <- meta[!rownames(meta) %in% c("Irrel_kd_3"), ]
```

## Check sample names match

```{r}
# Composition check
all(colnames(data) %in% rownames(meta)) # Should be TRUE
```

```{r}
# Order check
all(colnames(data) == rownames(meta))
```

```{r}
# If order check gives FALSE
# data_raw <- data
# data <- data_raw[, match(rownames(meta), colnames(data_raw))]
```

# Prefiltering (optional)

## Option 1: Filter genes where there are less than x counts

```{r}
# Filter genes where there are less than 10 counts
# idx <- rowSums(data) >= 10
# data <- data[idx,]
# data %>% dim() # [1] 17159     8
```

## Option 2: Filter genes where there are less than y samples having at least x counts

```{r}
# Filter genes where there are less than 3 samples having at least 5 counts
# idx <- rowSums(data >= 5) >= 3
# data <- data[idx,]
# data %>% dim() # [1] 15964     8
```

## Option 3: Only keep genes that have at least x count in at least the number of samples in the smallest group

```{r}
# Only keep genes that have at least 10 count in at least the number of samples in the smallest group
# smallest_group <- min(table(meta$sampletype))
# idx <- rowSums(data >= 10) >= smallest_group
# data <- data[idx,]
# data %>% dim() # [1] 15365     8
```

## Option 4: Only keep genes that have a total count of at least x in any condition

```{r}
# Only keep genes that have a total count of at least 10 in any condition
# cond <- meta$sampletype
# idx <- rep(FALSE, nrow(data))
# for (c in levels(cond)) {
#    idx <- idx | rowSums(data[, cond == c, drop=FALSE]) >= 10
# }
# data <- data[idx,]
# data %>% dim() # [1] 16409     8
```

## Option 5: Only keep genes that have at least x count in every sample of at least 1 condition

```{r}
# Only keep genes that have at least 10 count in every sample of at least 1 condition
# cond <- meta$sampletype
# idx <- rep(FALSE, nrow(data))
# for (c in levels(cond)) {
#  mat <- data[, cond == c, drop = FALSE]
#  idx <- idx | rowSums(mat >= 10) == ncol(mat)
# }
# data <- data[idx,]
# data %>% dim() # [1] 14901     8
```

## Option 6: Filtering after normalisation

```{r}
# Filter genes where there are less than 3 samples with normalized counts greater than or equal to 5
# dds <- DESeqDataSetFromMatrix(
#   countData = data,
#   colData = meta,
#   design = ~ sampletype
# )
# dds <- estimateSizeFactors(dds)
# idx <- rowSums(counts(dds, normalized=TRUE) >= 5 ) >= 3
# dds <- dds[idx,]
```

# QC

## Sample-level QC

### Count distribution

```{r}
# Count data long format
data_long <- data %>%
  gather(
    colnames(data),
    key = "sample_name",
    value = "raw_counts"
  )

# Count distribution
ggplot(
  data_long,
  aes(x = raw_counts)
) +
  geom_histogram(stat = "bin", bins = 100) +
  facet_wrap(~ sample_name) +
  xlab("Raw expression counts") +
  ylab("Number of genes") +
  ggtitle("Count distribution") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

### Mean - variance relation

```{r}
tmp <- data.frame(M = rowMeans(data), V = apply(data, 1, var))

ggplot(tmp, aes(x = M, y = V)) +
  geom_point(size = 1) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  scale_x_log10() + scale_y_log10() +
  geom_smooth(method = "gam", color = "blue") +
  xlab("Count mean") + ylab("Count variance") +
  ggtitle("Mean - variance relation",
          subtitle = "Red line = Poisson (mean = variance).") +
  theme_bw()
```

### Count normalisation

DESeq2-normalized counts: Median of ratios method

For more details: <https://hbctraining.github.io/Intro-to-DGE/lessons/02_DGE_count_normalization.html>

#### Create DESEq2 object

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = data,
  colData = meta,
  design = ~sampletype
)
```

```{r}
# Original count data
head(counts(dds))

# Original col data
colData(dds)
```

#### Generate size factors

```{r}
dds <- estimateSizeFactors(dds)
sizeFactors(dds)
```

```{r}
plot(
  sizeFactors(dds),
  main = "Size factor per sample",
  xlab = "Sample", ylab = "Size factor",
  cex.main = 1.2, pch = 16, col = "steelblue"
  )
abline(h = 1, col = "red", lty = 2)
```

#### Retrieve normalized counts

```{r}
normalized_counts <- counts(dds, normalized=TRUE)
```

```{r}
boxplot(
  log10(normalized_counts + 1),
  las = 2,
  main = "Normalized count distribution (log10)",
  ylab = "log10(normalized_counts + 1)",
  col = "lightblue", border = "grey40"
  )
```

```{r}
write.table(
  normalized_counts, 
  file = file.path(out_dir, "normalized_counts.tsv"), 
  sep = "\t", quote = FALSE, col.names = NA
 )
```

### Transformed normalized counts

Only for visualization in QC. Do not use for differential expression analysis

#### rlog transformation

rlog() works well on small datasets (n \< 30)

```{r}
rld <- rlog(dds, blind = TRUE)
rld_mat <- assay(rld)
```

```{r}
write.table(
  rld_mat, 
  file = file.path(out_dir, "normalized_counts_rlog.tsv"), 
  sep = "\t", quote = FALSE, col.names = NA
 )
```

#### vst transformation

vst() is faster than rlog(), recommended for medium to large datasets (n \> 30)

```{r}
vsd <- vst(dds, blind = TRUE) # n genes >= 1000
# vsd <- varianceStabilizingTransformation(dds, blind = TRUE) # n genes < 1000
vsd_mat <- assay(vsd)
```

```{r}
write.table(
  vsd_mat, 
  file = file.path(out_dir, "normalized_counts_vst.tsv"), 
  sep = "\t", quote = FALSE, col.names = NA
 )
```

#### log2(n+1) transformation

```{r}
ntd <- normTransform(dds)
ntd_mat <- assay(ntd)
```

```{r}
write.table(
  ntd_mat, 
  file = file.path(out_dir, "normalized_counts_log2.tsv"), 
  sep = "\t", quote = FALSE, col.names = NA
 )
```

### PCA

This example will use rlog transformation. vst or log2(n+1) could be used similarly

#### PC1 & PC2

```{r}
plotPCA(rld, intgroup = "sampletype") # Top 500 most variable genes
# plotPCA(rld, intgroup = "sampletype", ntop = 100) # Top 100 most variable genes
```

```{r}
# PCA data
pcaData <- plotPCA(rld, intgroup = c("sampletype", "MOVexpr"), returnData = TRUE)
# pcaData attributes
attributes(pcaData)
# Get the percentage variation
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
# Plot PCA using ggplot2
ggplot(
  pcaData,
  aes(
    x = PC1, y = PC2,
    color = sampletype, 
    shape = MOVexpr
    )
) +
  geom_point(size = 5) +
  labs(
    x = paste0("PC1 (", percentVar[1], "%)"),
    y = paste0("PC2 (", percentVar[2], "%)")
    ) +
    theme_classic()

ggsave(
  file = file.path(out_dir, "PCA.pdf"), 
  width = 8, height = 6, dpi = 300
  )
```

#### Additional PCs

```{r}
# PCs data
pca <- prcomp(t(rld_mat)) # all genes
# PCs attributes
attributes(pca)
# Build PCs dataframe
pca_df <- cbind(meta, pca$x)
pca_df
```

```{r}
# Percentage variation
percentVar <- round((pca$sdev^2) / sum(pca$sdev^2) * 100, 1)
percentVar
```

```{r}
# Plot PC3-4 using ggplot2
ggplot(
  pca_df,
  aes(
    x = PC3, y = PC4,
    color = sampletype, 
    shape = MOVexpr
    )
) +
  geom_point(size = 5) +
  labs(
    x = paste0("PC3 (", percentVar[3], "%)"),
    y = paste0("PC4 (", percentVar[4], "%)")
    ) +
  theme_classic()
```

#### Remove batch effect

Not relevant in this case. Just to show how to do it in case of confounding variable

```{r}
rld1 <- rld
assay(rld1) <- removeBatchEffect(assay(rld1), batch = colData(rld1)$MOVexpr) # Remove effect of MOVexpr
```

```{r}
plotPCA(rld1, intgroup = "sampletype") + 
  theme_bw() + 
  ggtitle("PCA per sampletype (after removing visual effect of MOVexpr)") 
```

Note: removeBatchEffect() only serves for visualisation. Statistical analysis should be correct using design argument in DESeqDataSetFromMatrix(): design = \~ confounder + condition (e.g. design = \~ sex + disease)

### Hierarchical Clustering

This example will use rlog transformation. vst or log2(n+1) could be used similarly

#### Correlation matrix

```{r}
# rlog matrix
rld_mat <- assay(rld)
```

```{r}
# Pairwise correlation
rld_cor <- cor(rld_mat)
```

```{r, fig.width=9, fig.height=6}
# Correlation heatmap, default
pheatmap(rld_cor, annotation = meta)
```

```{r, fig.width=9, fig.height=6}
# Correlation heatmap, customized
heat.colors <- brewer.pal(9, "Blues")
pheatmap(
  rld_cor,
  annotation = meta,
  color = heat.colors,
  cluster_cols = TRUE, cluster_rows = TRUE,
  border_color=NA, fontsize = 10, height=20
)
```

```{r}
# Save heatmap
# heat.colors <- brewer.pal(9, "Blues")
# png(
#   file = file.path(out_dir, "correlation_heatmap.png"),
#   width = 8, height = 8, units = "in", res = 300
#   )
# pheatmap(
#  rld_cor,
#  annotation = meta,
#  color = heat.colors,
#  cluster_cols = TRUE, cluster_rows = TRUE,
#  border_color=NA, fontsize = 10, height=20
# )
# dev.off()
```

#### Count matrix

```{r}
# Load normalized data
normalized_counts <- read.table(
  file = file.path(out_dir, "normalized_counts.tsv"),
  header = T, row.names = 1
  )
```

##### Genes highly expressed

```{r}
# Top 20 most expressed genes
top20_expression_genes <- order(rowMeans(as.matrix(normalized_counts)), decreasing = TRUE)[1:20]

normalized_counts[top20_expression_genes,] %>%
  row.names()
```

```{r}
pheatmap(
  rld_mat[top20_expression_genes, ],
  cluster_cols = TRUE, cluster_rows = TRUE,
  fontsize_row = 10,
  annotation_col = meta
)
```

##### Genes highly varied

```{r}
# Top 20 genes with most variance
top20_variance_genes <- order(rowVars(as.matrix(normalized_counts)), decreasing = TRUE)[1:20]

normalized_counts[top20_variance_genes,] %>%
  row.names()
```

```{r}
pheatmap(
  rld_mat[top20_variance_genes, ],
  cluster_cols = TRUE, cluster_rows = TRUE,
  fontsize_row = 10,
  annotation_col = meta
)
```

## Gene-level QC

Genes to omit:

-   Genes with zero counts in all samples

-   Genes with an extreme count outlier

-   Genes with a low mean normalized counts

DESeq2 will perform this filtering by default; however other DE tools, such as EdgeR will not

For more details: <https://hbctraining.github.io/Intro-to-DGE/lessons/05b_wald_test_results.html>

# Differential expression analysis: Pairwise comparison

## Run DESeq2

```{r}
# Create DESeqDataSet object
dds <- DESeqDataSetFromMatrix(
  countData = data,
  colData = meta,
  design = ~ sampletype
)
```

```{r}
# Set reference level
dds$sampletype <- relevel(dds$sampletype, ref = "control")
```

```{r}
dds <- DESeq(dds)
```

## Dispersion estimates

```{r, fig.width=16, fig.height=9}
par(mfrow = c(2,2))

dds_par <- estimateDispersions(dds, fitType = "parametric") # standard
dds_loc <- estimateDispersions(dds, fitType = "local") 
dds_mean <- estimateDispersions(dds, fitType = "mean")
# dds_gmp <- estimateDispersions(dds, fitType = "glmGamPoi") # single-cell, pseudo-bulk

plotDispEsts(dds_par, finalcol = NA, main = "parametric (default)")
plotDispEsts(dds_loc, finalcol = NA, main = "local")
plotDispEsts(dds_mean, finalcol = NA, main = "mean")
# plotDispEsts(dds_gmp, finalcol = NA, main = "glmGamPoi") 
```

```{r}
dds <- estimateDispersionsFit(dds, fitType = "parametric")
plotDispEsts(dds, main = "Final dispersion estimate (parametric)")
```

Examples of worrisome dispersion plots could be found at: <https://hbctraining.github.io/Intro-to-DGE/lessons/04b_DGE_DESeq2_analysis.html>

## Results

```{r}
# Define contrats: c("condition", "level_to_compare", "base_level")
contrast_oe <- c("sampletype", "MOV10_overexpression", "control")
contrast_kd <- c("sampletype", "MOV10_knockdown", "control")
```

```{r}
# ?results
```

### Overexpression vs control

#### Raw result

```{r}
# Extract using contrast
res_OE_unshrunken <- results(
  dds,
  contrast = contrast_oe,
  alpha = 0.05 # FDR adjusted p-value
)

res_OE_unshrunken
```

```{r}
# Result column info
mcols(res_OE_unshrunken, use.names = TRUE)
```

-   baseMean: mean of normalized counts for all samples

-   log2FoldChange: log2 fold change

-   lfcSE: standard error

-   stat: Wald statistic

-   pvalue: Wald test p-value

-   padj: BH adjusted p-values

```{r}
# Summarize results
summary(res_OE_unshrunken)
```

```{r}
resultsNames(dds)
```

```{r}
# Extract using name
res_OE_unshrunken_bis <- results(
  dds,
  name = "sampletype_MOV10_overexpression_vs_control",
  alpha = 0.05
)

res_OE_unshrunken_bis
```

```{r}
summary(res_OE_unshrunken_bis)
```

#### Shrinkage

Shrinkage of the LFC estimates toward zero when the information for a gene is low, which could include:

-   Low counts

-   High dispersion values

LFC shrinkage will not change the total number of genes that are identified as significantly differentially expressed

```{r}
resultsNames(dds)
```

```{r}
res_OE_norm <- lfcShrink(dds, coef="sampletype_MOV10_overexpression_vs_control", type="normal")
res_OE_apeglm <- lfcShrink(dds, coef="sampletype_MOV10_overexpression_vs_control", type="apeglm")
res_OE_ashr <- lfcShrink(dds, coef="sampletype_MOV10_overexpression_vs_control", type="ashr")
```

```{r, fig.width=16, fig.height=9}
par(mfrow = c(2, 2))
y.lim <- c(-2, 2)
DESeq2::plotMA(res_OE_unshrunken, ylim = y.lim, main = "No shrinkage")
DESeq2::plotMA(res_OE_norm, ylim = y.lim, main = "Shrinkage : normal")
DESeq2::plotMA(res_OE_apeglm, ylim = y.lim, main = "Shrinkage : apeglm")
DESeq2::plotMA(res_OE_ashr, ylim = y.lim, main = "Shrinkage : ashr")
```

```{r}
# Method chosen: apeglm
res_OE <- lfcShrink(
  dds, 
  coef="sampletype_MOV10_overexpression_vs_control",
  res = res_OE_unshrunken,
  type="apeglm")

summary(res_OE)
```

```{r}
# Convert to dataframe, sort by padj
res_OE_df <- res_OE %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  arrange(padj)
```

```{r}
write.table(
  res_OE_df, 
  file = file.path(out_dir, "Mov10oe_vs_control_result.tsv"), 
  sep = "\t", quote = FALSE, row.names = FALSE
 )
```

#### Significant DE genes

Use shrunken LFC

```{r}
# Set thresholds
padj.cutoff <- 0.05
lfc.cutoff <- log2(2)
```

```{r}
# DE genes
sig_OE_df <- res_OE_df %>%
  filter(padj < padj.cutoff & abs(log2FoldChange) > lfc.cutoff)
```

```{r}
head(sig_OE_df)
```

```{r}
write.table(
  sig_OE_df, 
  file = file.path(out_dir, "Mov10oe_vs_control_result_DE.tsv"), 
  sep = "\t", quote = FALSE, row.names = FALSE
 )
```

### KO vs control

```{r}
# Raw result
res_KD_unshrunken <- results(
  dds,
  contrast = contrast_kd,
  alpha = 0.05 # FDR adjusted p-value
)
```

```{r}
# Shrinkage: apeglm
res_KD <- lfcShrink(
  dds, 
  coef="sampletype_MOV10_knockdown_vs_control",
  res = res_KD_unshrunken,
  type="apeglm")

res_KD_df <- res_KD %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  arrange(padj)
```

```{r}
write.table(
  res_KD_df, 
  file = file.path(out_dir, "Mov10kd_vs_control_result.tsv"), 
  sep = "\t", quote = FALSE, row.names = FALSE
 )
```

```{r}
# Significant DE genes
padj.cutoff <- 0.05
lfc.cutoff <- log2(2)

sig_KD_df <- res_KD_df %>%
  filter(padj < padj.cutoff & abs(log2FoldChange) > lfc.cutoff)

head(sig_KD_df)
```

```{r}
write.table(
  sig_KD_df, 
  file = file.path(out_dir, "Mov10kd_vs_control_result_DE.tsv"), 
  sep = "\t", quote = FALSE, row.names = FALSE
 )
```

## Visualisation

Example: Overexpression vs control

```{r}
# Add row names to metadata
metadata <- meta %>%
  rownames_to_column("sample_name")
```

```{r}
# normalized_counts <- counts(dds, normalized = TRUE)
normalized_counts <- read.table(
  file.path(out_dir, "normalized_counts.tsv"),
  sep = "\t", header = T, row.names = 1
  )

normalized_counts <- normalized_counts %>%
  as.data.frame() %>%
  rownames_to_column("gene")
```

### p-value histogram

```{r}
hist(res_OE$pvalue, breaks = 100, probability = TRUE,
main = "p-value histogram", xlab = "p-value")
abline(h = 1, col = "red", lwd = 2)
```

Note: A p-value peak close to 0 suggests that there is a biological signal, i.e. there are DE genes

### Count plot

#### A single gene

##### plotCounts()

```{r}
plotCounts(dds, gene = "MOV10", intgroup = "sampletype")
```

##### ggplot2

```{r}
# Save plotCounts to a data frame object
MOV10_normCount <- plotCounts(dds, gene = "MOV10", intgroup = "sampletype", returnData = TRUE)
```


```{r}
# Plotting the MOV10 normalized counts
ggplot(
  MOV10_normCount,
  aes(x = sampletype, y = count, color = sampletype)
) +
  geom_point(position = position_jitter(width = 0.1), size = 3) +
  geom_text_repel(
    aes(label = rownames(MOV10_normCount))
  ) +
  labs(
    x = NULL, y = "Normalized counts", title = "MOV10"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```

#### Top DE genes

```{r}
# Top 20 DE genes by padj
top20_sig_OE_genes <- res_OE_df %>%
  arrange(padj) %>%
  pull(gene) %>%
  head(n=20)

top20_sig_OE_genes
```

```{r}
# Normalized counts for top 20 DE genes
top20_sig_OE_norm <- normalized_counts %>%
  filter(gene %in% top20_sig_OE_genes)

head(top20_sig_OE_norm)
```

```{r}
# Convert to long format
top20_sig_OE_norm <- top20_sig_OE_norm %>%
  gather(
    colnames(.)[-1],
    key = "sample_name",
    value = "normalized_counts"
  )

head(top20_sig_OE_norm)
```

```{r}
# Merge with metadata
top20_sig_OE_norm <- inner_join(metadata, top20_sig_OE_norm, by = "sample_name")

head(top20_sig_OE_norm)
```

```{r}
# Boxplots
ggplot(
  top20_sig_OE_norm,
  aes(x = gene, y = normalized_counts, color = sampletype)
) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  scale_y_log10() +
  scale_color_manual(
    labels = c("control", "MOV10_overexpression", "MOV10_knockdown"),
    values = c("gray", "red", "blue")
  ) +
  labs(
    x = NULL, y = "log10 normalized counts", title = "Top 20 Significant DE Genes"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

```{r}
# Facet boxplots
ggplot(
  top20_sig_OE_norm,
  aes(x = sampletype, y = normalized_counts, color = sampletype)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(height = 0, width = 0.15) +
  scale_y_continuous(trans = 'log10') +
  scale_color_manual(
    labels = c("control", "MOV10_overexpression", "MOV10_knockdown"),
    values = c("gray", "red", "blue")
  ) +
  ylab("log10 normalized expression level") +
  xlab("condition") +
  ggtitle("Top 20 Significant DE Genes in MOV10 overexpression vs Control") +
  facet_wrap(~ gene) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    plot.title = element_text(hjust = 0.5)
  )
```

```{r}
# Scatter plot
ggplot(
  top20_sig_OE_norm
  ) +
  geom_point(aes(x = gene, y = normalized_counts, color = sampletype)) +
  scale_y_log10() +
  xlab("Genes") +
  ylab("log10 Normalized Counts") +
  ggtitle("Top 20 Significant DE Genes") +
  theme_bw() +
	theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
	theme(plot.title = element_text(hjust = 0.5))
```

### Heatmap

```{r}
# Extract normalized expression for significant genes from the OE and control samples
sig_OE_norm <- normalized_counts[, c(1, 4:9)] %>%
  filter(gene %in% sig_OE_df$gene) %>%
  column_to_rownames("gene")
```

```{r}
# Set a color palette
heat_colors <- brewer.pal(6, "YlOrRd")

# Heatmap
pheatmap(
  sig_OE_norm,
  color = heat_colors,
  cluster_rows = TRUE, cluster_cols = TRUE,
  show_rownames = FALSE,
  annotation = meta,
  fontsize = 10,
  border_color = NA,
  scale = "row",
  fontsize_row = 10,
  height = 20
)
```

Note: Normalized gene counts are scaled by z-score standardization

### Volcano plot

```{r}
# Specify up/down/ns expressed genes
padj.cutoff <- 0.05
lfc.cutoff <- log2(2)
res_OE_df_annot <- res_OE_df %>%
  mutate(
    expression = case_when(
      padj < padj.cutoff & log2FoldChange > lfc.cutoff ~ "up",
      padj < padj.cutoff & log2FoldChange < -lfc.cutoff ~ "down",
      TRUE ~ "ns"
    )
  )
```

```{r}
# Data range
range(res_OE_df_annot$log2FoldChange, na.rm = TRUE) # x axis
max(-log10(range(res_OE_df_annot$padj, na.rm = TRUE))) # y axis
```


```{r}
# Volcano plot
res_OE_df_annot %>%
  ggplot(
    aes(x = log2FoldChange, y = -log10(padj), color = expression, alpha = 0.5)
  ) +
  geom_point(size = 3) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed"
  ) +
  geom_vline(
    xintercept = c(-lfc.cutoff, lfc.cutoff),
    linetype = "dashed"
  ) +
  scale_color_manual(
    breaks = c("up", "down", "ns"),
    values = c("tomato", "steelblue", "grey")
  ) +
  guides(alpha = "none") +
  coord_cartesian(xlim = c(-6, 6), ylim = c(0, 200)) +
  labs(
    title = "Gene expression changes between MOV10 overexpression vs Control",
    x = "log2(FC)", y = "-log10(padj)", color = "Expression\nchange"
  ) +
  theme_classic()
```

```{r}
# Select genes to label
res_OE_df_annot %>%
  mutate(abs_log2FoldChange = abs(log2FoldChange)) %>%
  filter(-log10(padj) > 50 | abs(log2FoldChange) > 5) %>%
  dim()
```

```{r}
# Label genes dataframe
res_OE_df_annot_label <- res_OE_df_annot %>%
  mutate(abs_log2FoldChange = abs(log2FoldChange)) %>%
  filter(-log10(padj) > 50 | abs(log2FoldChange) > 5) %>%
  mutate(label = gene)
```

```{r}
# Volcano plot with top genes labeling
ggplot(
  data = res_OE_df_annot,
  aes(x = log2FoldChange, 
      y = -log10(padj))
  ) +
  geom_point(
    aes(colour = expression),
    alpha = 0.6,
    shape = 19,
    size = 2
    ) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed"
  ) + 
  geom_vline(
    xintercept = c(-lfc.cutoff, lfc.cutoff),
    linetype = "dashed"
  ) +
  geom_label_repel(
    data = res_OE_df_annot_label,
    aes(label = label),
    force = 2
  ) +
  scale_color_manual(
    breaks = c("up", "down", "ns"),
    values = c("darkred", "darkblue", "grey")
  ) +
  scale_x_continuous(
    breaks = c(seq(-8, 8, 2)),
    limits = c(-8, 8)
    ) +
  coord_cartesian(ylim = c(0, 200)) +
  labs(
    title = "Gene expression changes between MOV10 overexpression vs Control",
    x = "log2(FC)", 
    y = "-log10(padj)", 
    color = "Expression\nchange"
  ) +
  theme_light()
```

# Differential expression analysis: Likelihood ratio test (LRT)

LRT is an alternative solution when evaluating expression change across more than two levels.

Genes which are identified as significant, are those that are changing in expression in any direction across the different factor levels.

log2FoldChange in LRT test is an arbitrary coefficient and should not be used

For more details: <https://hbctraining.github.io/Intro-to-DGE/lessons/05a_hypothesis_testing.html#likelihood-ratio-test-lrt>

## LRT

```{r}
dds <- DESeqDataSetFromMatrix(
    countData = data,
    colData = meta,
    design = ~ sampletype
)

# Likelihood ratio test
dds_lrt <- DESeq(dds, test = "LRT", reduced = ~ 1)
```

```{r}
# LRT result
res_lrt <- results(
  dds_lrt, 
  alpha = 0.05 # FDR adjusted p-value
  )
res_lrt
```

```{r}
# LRT result summary
summary(res_lrt)
```

```{r}
# Convert to dataframe, sort by padj
res_lrt_df <- res_lrt %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  arrange(padj)
```

```{r}
write.table(
  res_OE_df, 
  file = file.path(out_dir, "lrt_result.tsv"), 
  sep = "\t", quote = FALSE, row.names = FALSE
 )
```

```{r}
# Set thresholds
padj.cutoff <- 0.05

# DE genes
sig_lrt_df <- res_lrt_df %>%
  filter(padj < padj.cutoff)

head(sig_lrt_df)
```

```{r}
write.table(
  sig_lrt_df, 
  file = file.path(out_dir, "lrt_result_DE.tsv"), 
  sep = "\t", quote = FALSE, row.names = FALSE
 )
```

```{r}
nrow(sig_lrt_df)
nrow(sig_OE_df)
nrow(sig_KD_df)
```

Note: The number of significant genes observed from the LRT is quite high. To reduce the number of significant genes, we can increase the stringency of our FDR threshold

## Significant DE genes clustering

```{r}
# Significant DE gene list, padj cutoff 1e-7
sig_lrt_genes <- sig_lrt_df %>%
  filter(padj < 1e-7) %>%
  pull(gene)
```

```{r}
# Load rlog data
rld_mat <- read.table(
  file = file.path(out_dir, "normalized_counts_rlog.tsv"),
  header = T, row.names = 1
  )
```

```{r}
# Subset rlog data for significant DE genes
rld_sig_lrt_mat <- rld_mat[sig_lrt_genes, ]
```

```{r}
# DE genes clustering
clusters <- degPatterns(
   rld_sig_lrt_mat,
   metadata = meta,
   time = "sampletype",
   col = NULL
)

# Rotate x-axis labels
clusters$plot +
  theme(
    axis.text.x = element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1
    )
  )
```

```{r}
# class(clusters)
# clusters$df
# Number of genes per cluster
clusters$df %>%
  count(cluster)
```

```{r}
# Example: Extract genes in cluster 1 where KD ~ CTL < OE
cluster_1_genes <- clusters$df %>%
  filter(cluster == 1) %>%
  pull(genes)

length(cluster_1_genes)
```

# Functional analysis: Over-representation analysis (ORA)

Example: Overexpression vs control, GO terms

## ORA

```{r}
# Load DEA data
res_OE_df <- read.table(
  file = file.path(out_dir, "Mov10oe_vs_control_result.tsv"),
  header = T
  )

res_OE_df %>%
  dim
```

```{r}
# Filter out untested genes, i.e. genes with padj = NA
res_OE_noNA_df <- res_OE_df %>%
  filter(padj != "NA")

res_OE_noNA_df %>%
  dim
```

```{r}
# Background genes: All genes tested for significance
all_OE_genes <- as.character(res_OE_noNA_df$gene)

length(all_OE_genes)
```

```{r}
# Significant DE genes
padj.cutoff <- 0.05

sig_OE_genes <- res_OE_noNA_df %>%
  filter(padj < padj.cutoff) %>%
  pull(gene)

length(sig_OE_genes)
```

```{r}
# ORA analysis
ora_res <- enrichGO(
  gene = sig_OE_genes,     # Gene list of interest
  universe = all_OE_genes, # Background genes
  OrgDb = org.Hs.eg.db,    # Organism database
  keyType = "SYMBOL",      # Key type of gene identifiers in gene list 
  ont = "BP",              # Oncology: BP (Biological Process)
  pAdjustMethod = "BH",    # Multiple testing correction
  pvalueCutoff  = 0.05,    # P-value cutoff
  qvalueCutoff = 0.05      # Q-value cutoff
  )
```

Note: Parameters to consider

- DE thresholds: padj, log2FoldChange

- Gene list: DE genes, Up, Down

- Ontology: "BP", "MF", "CC" or "ALL"

- Ontology pAdjustMethod: "BH", "bonferroni", ... (?enrichGO for more details)

- Ontology pvalueCutoff, qvalueCutoff

```{r}
# ORA result table
ora_res_df <- data.frame(ora_res)

head(ora_res_df)
```

```{r}
write.table(
  ora_res_df,
  file = file.path(out_dir, "Mov10oe_vs_control_result_DE_ORA_BP.tsv"),
  sep = "\t", quote = FALSE, row.names = FALSE
  )
```

```{r}
range(ora_res_df$p.adjust, na.rm = TRUE)
```

## Visualisation

### Dotplot

```{r, fig.width=9, fig.height=16}
# Display gene ratio: Nb of genes related to GO term / total number of significant genes
dotplot(
  ora_res, 
  showCategory=30,
  title="GO Biological Processes Enrichment in DE genes"
  )
```

### Barplot

```{r, fig.width=9, fig.height=16}
# Display gene count
barplot(
  ora_res, 
  showCategory = 30, 
  title="GO Biological Processes Enrichment in DE genes"
  )
```

### Enrichment GO map

```{r}
# Add similarity matrix to the termsim slot of enrichment result
ora_res_sim <- enrichplot::pairwise_termsim(ora_res)
```

```{r, fig.width=16, fig.height=16}
# Enrichmap clusters the 50 most significant (by padj) GO terms to visualize relationships between terms
emapplot(ora_res_sim, showCategory = 50)
```

### Category netplot

```{r}
# Extract the log2FoldChange values
res_OE_LFC <- res_OE_noNA_df$log2FoldChange
names(res_OE_LFC) <- res_OE_noNA_df$gene
head(res_OE_LFC)
```

#### Top significant GO terms 

```{r, fig.width=16, fig.height=16}
# Cnetplot: genes associated with top 5 significant terms
cnetplot(
  ora_res, 
  showCategory = 5,
  foldChange = res_OE_LFC
  )
```

```{r, fig.width=16, fig.height=16}
# Set LFC threshold to narrow down the color scale
res_OE_LFC_cutoff <- ifelse(res_OE_LFC > 2, 2, res_OE_LFC)
res_OE_LFC_cutoff <- ifelse(res_OE_LFC < -2, -2, res_OE_LFC)

# Set threshold for absolute fold change to filter items
cnetplot(
  ora_res, 
  showCategory = 5,
  foldChange = res_OE_LFC_cutoff,
  fc_threshold = 0.5
  )
```

#### Selected GO terms

```{r}
ora_res_df$Description
```


```{r, fig.width=16, fig.height=16}
# E.g. "RNA splicing", "RNA localization", "RNA transport"
ora_res_bis <- ora_res
ora_res_bis@result <- ora_res@result[c(6,7,14),]
cnetplot(
  ora_res_bis, 
  showCategory = 3,
  foldChange = res_OE_LFC
  )
```

# Functional analysis: Gene set enrichment analysis (GSEA)

Example: Overexpression vs control, KEGG pathways

```{r}
# Load DEA data
res_OE_df <- read.table(
  file = file.path(out_dir, "Mov10oe_vs_control_result.tsv"),
  header = T
  )

res_OE_df %>%
  dim
```

## Data preparation

```{r}
# Retrieve Entrez IDs
entrez <- mapIds(
  org.Hs.eg.db,
  keys = res_OE_df$gene,
  column = "ENTREZID",
  keytype = "SYMBOL",
  multiVals = "first"
)

# Convert to dataframe
entrez_df <- enframe(
  entrez,
  name = "SYMBOL",
  value = "ENTREZID"
)

head(entrez_df)
```

```{r}
# Add Entrez IDs to result table
colnames(entrez_df) <- c("gene", "entrezid")
res_OE_entrez_df <- inner_join(
  res_OE_df, entrez_df, by = "gene"
  )
```

```{r}
# Remove NA values
res_OE_entrez_df <- res_OE_entrez_df %>%
  filter(!is.na(entrezid))
```

```{r}
# Remove Entrez duplucates
res_OE_entrez_df <- res_OE_entrez_df[which(duplicated(res_OE_entrez_df$entrezid) == F), ]
```

```{r}
# Extract the log2FoldChange values
res_OE_LFC <- res_OE_entrez_df$log2FoldChange
names(res_OE_LFC) <- res_OE_entrez_df$entrezid

head(res_OE_LFC)
```

```{r}
# Sort the log2FoldChange values in decreasing order
res_OE_LFC_sorted <- sort(res_OE_LFC, decreasing = TRUE)

head(res_OE_LFC_sorted)
```

## GSEA

```{r}
set.seed(42) # reproducibility
```

```{r}
# GSEA using gene sets from KEGG pathways
gsea_kegg <- gseKEGG(
  geneList = res_OE_LFC_sorted,
  organism = "hsa",
  minGSSize = 20, # minimum gene set size
  pvalueCutoff = 0.05, 
  verbose = FALSE
  )
```

Note: The organisms with available KEGG pathway information are listed in <https://www.genome.jp/kegg/tables/br08606.html>

```{r}
# GSEA results
gsea_kegg_res <- gsea_kegg@result

head(gsea_kegg_res)
```

```{r}
write.table(
  gsea_kegg_res,
  file = file.path(out_dir, "Mov10oe_vs_control_result_DE_GSEA_KEGG.tsv"),
  sep = "\t", quote = FALSE, row.names = FALSE
  )
```

## Visualisation

```{r, fig.width=9, fig.height=12}
# GSEA plot for a single enriched pathway, e.g. hsa04060
gseaplot(gsea_kegg, geneSetID = 'hsa04060') # up-regulation
```

```{r}
# GSEA plot for a single enriched pathway
gseaplot(gsea_kegg, geneSetID = 'hsa03008') # down-regulation
```

```{r}
# KEGG pathway map
pathview(
  gene.data = res_OE_LFC_sorted,
  pathway.id = "hsa03008",
  species = "hsa",
  limit = list(
    gene = 2, # value gives the max/min limit for foldchanges
    cpd = 1
    )
  )

# Move output files
files <- list.files(
  path = ".",
  pattern = "^hsa03008",
  full.names = TRUE
)
file.rename(files, file.path(out_dir, basename(files)))
```

```{r}
knitr::include_graphics(
  file.path(out_dir, "hsa03008.pathview.png"),
  )
```

# References

<https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html>

<https://hbctraining.github.io/Intro-to-DGE/schedule/links-to-lessons.html>
